---
title: "Chapter 1 — Reference panel overview"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
})

read_tsv_safe <- function(p, ...) if (file.exists(p)) read_tsv(p, show_col_types = FALSE, ...) else tibble()

# `*_lengths.tsv` is the seqkit fx2tab output — two cols, no header.
mit_lens   <- read_tsv_safe("../outputs/qc/mit_lengths.tsv",
                            col_names = c("id","len"))
s18_lens   <- read_tsv_safe("../outputs/qc/18S_lengths.tsv",
                            col_names = c("id","len"))
mit_class  <- read_tsv_safe("../outputs/qc/mit_length_classification.tsv")
s18_class  <- read_tsv_safe("../outputs/qc/18S_length_classification.tsv")
s18_cov    <- read_tsv_safe("../outputs/cross_species/18S_species_coverage.tsv")

# MIT IDs are self-describing — species is the prefix `P<species>_…`.
# Strip the leading `P` to canonicalise.
mit_per_species <- mit_lens |>
  mutate(species = str_to_lower(str_extract(id, "^P[A-Za-z]+"))) |>
  mutate(species = str_replace(species, "^p", "")) |>
  count(species, name = "n_sequences", sort = TRUE)
```

The two reference panels are the inputs the entire pipeline operates on.
Both come straight from NCBI multi-fasta DBs and are filtered to the 11
canonical target species before any cross-species or alignment work.

## Reference panels

```{r panel-counts}
panel <- tibble(
  marker = c("MIT", "18S"),
  n_sequences = c(nrow(mit_lens), nrow(s18_lens)),
  median_len  = c(median(mit_lens$len), median(s18_lens$len)),
  n_partial   = c(sum(mit_class$is_partial, na.rm = TRUE),
                  sum(s18_class$is_partial, na.rm = TRUE))
)
kable(panel, col.names = c("Marker", "Sequences (target spp.)",
                           "Median length (bp)", "Partial (<90% of max)"))
```

"Partial" is `len < 0.9 * max(len)` per marker. MIT mitogenomes cluster
tightly around the full-genome length (~6 kb); 18S has a wider spread
because public 18S sequences vary from short PCR fragments to full-gene
inserts.

## MIT references per species

```{r mit-per-species}
kable(mit_per_species, col.names = c("Species", "MIT references"))
```

## 18S coverage check

The 18S species-coverage check is a v1 acceptance criterion (HANDOFF.md
Step 4). Every canonical target species needs at least one 18S
reference for that pair to be evaluable on 18S; species with zero
references are flagged here so downstream resolution-table NAs are
attributable.

```{r s18-coverage}
kable(s18_cov, col.names = c("Species", "18S references", "Covered (≥1)"))

n_covered <- sum(s18_cov$is_covered, na.rm = TRUE)
n_total   <- nrow(s18_cov)
missing   <- s18_cov |> filter(!is_covered) |> pull(species)
```

`r n_covered`/`r n_total` target species are covered on 18S. Missing:
**`r if (length(missing) == 0) "none" else paste(missing, collapse = ", ")`**.
Any pair involving a missing species is `NA` on the 18S column of the
resolution table; for those pairs the assay's resolution claim depends
on MIT alone.

## Methods

Reference staging and length tables: [`workflow/rules/01_blast.README.md`](../workflow/rules/01_blast.README.md),
[`workflow/rules/02_cross_species.README.md`](../workflow/rules/02_cross_species.README.md).
Coverage check: [`workflow/rules/04_resolution.README.md`](../workflow/rules/04_resolution.README.md).
