---
title: "Chapter 3 — 18S cross-species similarity"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({ library(tidyverse); library(knitr) })
read_tsv_safe <- function(p) if (file.exists(p)) read_tsv(p, show_col_types = FALSE) else tibble()

s18_blast <- read_tsv_safe("../outputs/blast/self_blast_18S.tsv")
s18_susp  <- read_tsv_safe("../outputs/cross_species/18S_suspicious.tsv")
s18_loose <- read_tsv_safe("../outputs/cross_species/18S_suspicious_loose.tsv")
s18_pair  <- read_tsv_safe("../outputs/cross_species/18S_pair_summary.tsv")
s18_cov   <- read_tsv_safe("../outputs/cross_species/18S_species_coverage.tsv")
s18_clos  <- read_tsv_safe("../outputs/cross_species/18S_closest_per_seq.tsv")

n_blast   <- nrow(s18_blast)
n_susp_st <- nrow(s18_susp)
n_susp_lo <- nrow(s18_loose)
n_pairs   <- nrow(s18_pair)
n_strict  <- sum(s18_pair$strict_ambiguity, na.rm = TRUE)
```

18S takes the same self-BLAST → species-extraction → top-5 HSPs scaffold
as MIT, with one extra step: 18S headers carry descriptive species
names rather than self-describing IDs, so an explicit
`outputs/cross_species/18S_acc_to_header.tsv` mapping table parses
species out of headers before the cross-species logic runs.

`r format(n_blast, big.mark = ",")` BLAST hits in `self_blast_18S.tsv`.
The strict filter
(`pident ≥ 99 AND cov_shorter ≥ 0.99 AND shorter_len ≥ 10`) returns
**`r n_susp_st`** hits.

## Strict filter — empty

The strict filter file is header-only. At pident ≥ 99 with full
shorter-sequence coverage, **no cross-species 18S pair is flagged.**
This is the headline result that supports the dual-marker rationale in
the resolving direction: 18S separates the *vivax*/*simium* pair MIT
cannot resolve (Chapter 2).

The strict filter being too aggressive to surface partial ambiguities
is the next problem. The looser ladder addresses it.

## Looser-threshold re-pass

The looser filter (`pident ≥ 98 AND mincov ≥ 0.85`) is applied to the
same `self_blast_18S.tsv` — no re-BLAST. *mincov* = `min(qcov, scov)`
is more conservative than *cov_shorter* (always ≤ cov_shorter), so this
ladder is genuinely looser-on-pident-and-stricter-on-coverage relative
to the strict filter.

```{r s18-loose-table}
if (nrow(s18_loose) > 0) {
  s18_loose |>
    select(query_species, subject_species, pident, aln_len,
           qcov, scov, mincov, cov_shorter, mismatch, bitscore) |>
    arrange(desc(pident)) |>
    kable(digits = c(0, 0, 3, 0, 4, 4, 4, 4, 0, 1))
} else {
  cat("(empty — no pair surfaces at the looser threshold)\n")
}

n_unique_pairs_lo <- if (nrow(s18_loose) > 0) {
  s18_loose |>
    transmute(p = pmap_chr(list(query_species, subject_species),
                           \(a, b) paste(sort(c(a, b)), collapse = " vs "))) |>
    distinct() |>
    nrow()
} else 0L
```

`r n_susp_lo` HSPs surface across `r n_unique_pairs_lo` unordered
species pair(s). The headline finding is `cynomolgi` / `fieldi` —
partial 18S ambiguity that the strict filter is too aggressive to
catch. The report should treat these as "watch this pair" rather than
"unresolvable", but the strict-version claim ("18S resolves all panel
pairs") needs the looser-version footnote attached.

## Per-pair summary

```{r s18-pair-summary}
s18_pair |>
  arrange(desc(max_mincov)) |>
  head(15) |>
  kable(digits = c(NA, 0, 3, 0, 4, 4, 0, 0, NA, NA),
        col.names = c("Pair", "n_HSPs", "best_pident", "max_aln_len",
                      "max_cov_shorter", "max_mincov", "min_mismatch",
                      "min_gapopen", "strict_ambig.", "ladder_class"))
```

`r n_pairs` cross-species pairs across 18S. `r n_strict` strict-
ambiguity flag(s). The ladder column shows where the partial
ambiguities sit — pairs at `high_confidence_possible_ambiguity` are
the ones the looser filter surfaces.

## Coverage caveat

```{r s18-coverage}
kable(s18_cov, col.names = c("Species", "18S references", "Covered"))

missing <- s18_cov |> filter(!is_covered) |> pull(species)
n_uncov <- length(missing)
```

`r n_uncov` panel species has zero 18S references in the panel
(**`r if (n_uncov == 0) "none" else paste(missing, collapse = ", ")`**).
For *that* species, 18S can't say anything; the resolution table
correctly returns NA for every pair involving it. Several other species
have only 1 reference — see Chapter 5 for the per-pair confidence
notes.

## Methods

Filter implementation:
[`workflow/rules/02_cross_species.README.md`](../workflow/rules/02_cross_species.README.md).
Looser-threshold re-pass and species-coverage check:
[`workflow/rules/04_resolution.README.md`](../workflow/rules/04_resolution.README.md).
