---
title: "Chapter 4 — Alignment, entropy, and primer-candidate windows"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({ library(tidyverse); library(knitr) })
read_tsv_safe <- function(p) if (file.exists(p)) read_tsv(p, show_col_types = FALSE) else tibble()

mit_pp     <- read_tsv_safe("../outputs/entropy/mit.per_position.tsv")
s18_pp     <- read_tsv_safe("../outputs/entropy/18S.per_position.tsv")
mit_core   <- read_tsv_safe("../outputs/entropy/mit.core_bounds.tsv")
s18_core   <- read_tsv_safe("../outputs/entropy/18S.core_bounds.tsv")
mit_pc     <- read_tsv_safe("../outputs/entropy/mit.primer_candidates.tsv")
s18_pc     <- read_tsv_safe("../outputs/entropy/18S.primer_candidates.tsv")

mit_aln_len <- nrow(mit_pp)
s18_aln_len <- nrow(s18_pp)

core_one <- function(d) {
  if (nrow(d) == 0) return(c(NA_integer_, NA_integer_, NA_integer_))
  s <- as.integer(d$core_left[1]); e <- as.integer(d$core_right[1])
  c(s, e, e - s + 1L)
}
mit_c <- core_one(mit_core)
s18_c <- core_one(s18_core)
```

The target-species subsets are aligned with `mafft --auto`. Per-position
Shannon entropy is computed over A/C/G/T (gaps ignored), with coverage
defined as the fraction of sequences having a non-gap base at that
position. Primer-candidate windows are sliding-window slices that pass
low-entropy and high-coverage thresholds at one end of the alignment
("left" = 5′ in alignment coordinates, "right" = 3′).

## Alignment dimensions

```{r aln-dims}
tibble(marker = c("MIT", "18S"),
       n_seqs = c(0, 0),  # filled below from the alignment file directly
       aln_len = c(mit_aln_len, s18_aln_len)) |>
  mutate(n_seqs = c(
    length(grep("^>", readLines("../outputs/alignment/ma_mit.target.fasta"))),
    length(grep("^>", readLines("../outputs/alignment/ma_18S.target.fasta")))
  )) |>
  kable(col.names = c("Marker", "Sequences", "Alignment length (bp)"))
```

## Core bounds

The "core" of each alignment is the contiguous region with coverage
≥ 90% and a minimum 50-bp run (`config.yaml::entropy.cov_thr` and
`min_run`). It's where primer candidates are searchable; outside the
core, gappy reference sequences make the entropy signal noisy.

```{r core-bounds}
tibble(marker = c("MIT", "18S"),
       core_start = c(mit_c[1], s18_c[1]),
       core_end   = c(mit_c[2], s18_c[2]),
       core_len   = c(mit_c[3], s18_c[3])) |>
  kable(col.names = c("Marker", "Core start", "Core end", "Core length"))
```

For comparison, the legacy run reports MIT core 749→6656 (length 5908)
and 18S core 317→2218 (length 1902). The reproduction here matches.

## Entropy + coverage panels

Both panels share the same x-axis (alignment position) and overlay
candidate primer windows (darker shade) and internal high-entropy
regions (lighter shade) on the per-position curve.

::: {#fig-mit-entropy layout-ncol=1}

![MIT entropy vs alignment position. Low-entropy plateaus inside the
core are where conserved primer-binding sites can sit; the high-entropy
spikes are where the assay's discriminating power lives at sequencing
time.](figures/mit_entropy_vs_position.png){#fig-mit-entropy-png}

![MIT coverage vs alignment position. The 1.0 plateau across the core
means every reference contributes at every position there; coverage
drops at the alignment ends are gappy/truncated references and are why
the core bounds clip them off.](figures/mit_coverage_vs_position.png){#fig-mit-coverage-png}

MIT panel: (a) entropy, (b) coverage, both vs alignment position.
:::

::: {#fig-18s-entropy layout-ncol=1}

![18S entropy vs alignment position. 18S shows alternating conserved
and variable blocks (the "A-type / S-type" structure described in the
published reference panel — see Chapter 6).](figures/18S_entropy_vs_position.png){#fig-18s-entropy-png}

![18S coverage vs alignment position. The shorter alignment plus more
heterogeneous reference lengths give a coverage profile that's flatter
in the core and tapers more steeply at both ends than MIT.](figures/18S_coverage_vs_position.png){#fig-18s-coverage-png}

18S panel: (a) entropy, (b) coverage.
:::

## Primer-candidate windows

```{r primer-candidates}
tibble(marker = c("MIT", "18S"),
       n_candidate_windows = c(nrow(mit_pc), nrow(s18_pc)),
       left  = c(sum(mit_pc$end_region == "left", na.rm = TRUE),
                 sum(s18_pc$end_region == "left", na.rm = TRUE)),
       right = c(sum(mit_pc$end_region == "right", na.rm = TRUE),
                 sum(s18_pc$end_region == "right", na.rm = TRUE))) |>
  kable(col.names = c("Marker", "Total windows", "Left-end", "Right-end"))
```

Each row of `<marker>.primer_candidates.tsv` is one window position.
The `end_region` column flags which end of the core the window falls
near (where a primer would actually be useful). Phase 5 work — primer3
selection inside these windows, multiplex Tm/dimer checks, in-silico
specificity — is deferred per HANDOFF.md.

## Methods

Alignment, entropy, and plotting:
[`workflow/rules/03_alignment.README.md`](../workflow/rules/03_alignment.README.md).
The entropy implementation lives at
`scripts/python/seq_entropy.py` (copied unchanged from the legacy
`scripts/legacy/speciation_long/scripts/seq_entropy.py`); the plotting
script is `scripts/R/plot_entropy.R` (argv-driven refactor of the
legacy version).
