---
title: "Chapter 5 — Cross-marker resolution table"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({ library(tidyverse); library(knitr) })
read_tsv_safe <- function(p) if (file.exists(p)) read_tsv(p, show_col_types = FALSE) else tibble()

res <- read_tsv_safe("../outputs/cross_species/resolution_table.tsv")

# Coerce empty strings (NA on disk) to logical NA explicitly
to_lgl <- function(x) {
  out <- as.logical(toupper(x))
  out
}
if (nrow(res) > 0) {
  for (c in c("mit_resolved","18S_resolved","18S_resolved_loose","combined_resolved")) {
    res[[c]] <- to_lgl(res[[c]])
  }
}

n_total <- nrow(res)
tot <- function(col) {
  list(T  = sum(res[[col]] %in% TRUE),
       F_ = sum(res[[col]] %in% FALSE),
       NA_= sum(is.na(res[[col]])))
}
mit  <- tot("mit_resolved")
s18  <- tot("18S_resolved")
loo  <- tot("18S_resolved_loose")
comb <- tot("combined_resolved")
```

The resolution table is the v1 deliverable. One row per unordered panel
pair. Three independent resolution columns (`mit_resolved`,
`18S_resolved`, `18S_resolved_loose`) and a derived `combined_resolved`
folded over MIT and the *strict* 18S column.

`r n_total` panel pairs total
(`r format(n_total, big.mark = ",")` = 11 × 10 / 2).

## Headline counts

```{r counts-table}
tibble(
  Column = c("mit_resolved", "18S_resolved (strict)", "18S_resolved_loose",
             "combined_resolved"),
  TRUE_  = c(mit$T, s18$T, loo$T, comb$T),
  FALSE_ = c(mit$F_, s18$F_, loo$F_, comb$F_),
  NA_    = c(mit$NA_, s18$NA_, loo$NA_, comb$NA_)
) |> kable(col.names = c("Column", "TRUE", "FALSE", "NA"))
```

**Reading the row.** TRUE means the pair is resolvable (the species
have at least one reference each on that marker, and the pair does not
appear in that marker's suspicious table). FALSE means the pair *is*
in the suspicious table. NA means at least one species has zero
references on that marker — the assay can't say either way. The
combined column uses Kleene-style OR over MIT and strict-18S: any TRUE
collapses to TRUE; mixed FALSE/NA collapses to FALSE; only NA-on-both
yields NA.

The single MIT FALSE is `simium vs vivax`, exactly as expected from
the legacy outputs. Every panel pair is `combined_resolved = TRUE`.

## Borderline rows

```{r borderline}
res_disp <- res |>
  rename(`MIT` = mit_resolved,
         `18S strict` = `18S_resolved`,
         `18S loose`  = `18S_resolved_loose`,
         `Combined`   = combined_resolved) |>
  mutate(across(c(`MIT`, `18S strict`, `18S loose`, `Combined`),
                ~ recode(as.character(.x), "TRUE" = "T", "FALSE" = "F", .missing = "NA")))

# Anything that's not all-T (or has notes attached) is "borderline" worth surfacing
borderline <- res_disp |> filter(`MIT` != "T" | `18S strict` != "T" |
                                 `18S loose` != "T" | `Combined` != "T" |
                                 nchar(notes) > 0)
borderline |>
  kable()
```

The `notes` column carries hard-coded annotations the human reader
benefits from at a glance: the *vivax*/*simium* MIT ambiguity reference
to *P. simium* zoonotic origin, and the "18S coverage low" warning that
fires whenever either species has ≤ 1 18S reference.

## Validation

Per `HANDOFF.md` Step 4, the validation gate is that the resolution
table reproduces the *vivax*/*simium* MIT ambiguity finding from the
legacy outputs. If `simium vs vivax` ever shows `mit_resolved = TRUE`,
something is wrong upstream.

```{r validation}
vs <- res |> filter(pair == "simium vs vivax")
if (nrow(vs) == 0) cat("ERROR: simium vs vivax row missing!\n")
vs |> kable()
```

Pass: `mit_resolved = FALSE`, `18S_resolved = TRUE`, `combined_resolved = TRUE`.

## What this says about the assay

- **MIT alone** has one residual ambiguity (*vivax*/*simium*) the panel
  cannot resolve at high confidence. Calls into samples where simian
  spillover into humans is plausible should not be made on MIT alone.
- **18S strict** resolves every panel pair, but the 18S coverage gap
  (zero references for *simiovale*) means 10/55 pairs are NA on 18S.
  For those pairs the panel rests on MIT.
- **18S looser** is a hedge: a meaningful but not-strict signal on
  *cynomolgi*/*fieldi* the strict filter doesn't catch. The report
  treats this as "watch this pair" rather than "unresolvable".
- **Combined**, every pair is resolvable. The dual-marker rationale
  holds at v1.

## Methods

Resolution table generation:
[`workflow/rules/04_resolution.README.md`](../workflow/rules/04_resolution.README.md).
