---
title: "Chapter 6 — Published-panel comparison (Imwong et al., 2018)"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  # Biostrings first — its S4Vectors::rename / S4Vectors::filter would
  # otherwise mask dplyr's. Loading tidyverse last keeps the dplyr verbs
  # in scope for the rest of the chunk.
  library(Biostrings)
  library(officer)
  library(knitr)
  library(tidyverse)
})

read_tsv_safe <- function(p) if (file.exists(p)) read_tsv(p, show_col_types = FALSE) else tibble()

docx_path <- "../jiy519_suppl_supplementary-table1-5 (2).docx"
mit_pc    <- read_tsv_safe("../outputs/entropy/mit.primer_candidates.tsv")
s18_pc    <- read_tsv_safe("../outputs/entropy/18S.primer_candidates.tsv")
s18_aln   <- if (file.exists("../outputs/alignment/ma_18S.target.fasta"))
               readDNAStringSet("../outputs/alignment/ma_18S.target.fasta") else NULL
```

The Imwong et al. (2018) supplementary table —
`jiy519_suppl_supplementary-table1-5 (2).docx` at the project root — is
the published *Plasmodium* primer reference panel this assay aligns
against. This chapter pulls the primer sequences out of the docx and
checks where they sit on our 18S alignment, and which species the
published panel covers vs. ours.

## What the docx contains

Read with `officer::read_docx()` and a regex sweep for DNA-string cells.

```{r docx-read}
docx <- read_docx(docx_path)
sx   <- docx_summary(docx)

dna_re <- "^[ACGTRYSWKMBDHVN]{15,40}$"

# In the docx, Supplementary Table 2 lays primers out as
#   col 1: round (Nest1 / Nest2 / Sequencing)
#   col 2: primer name (PlasmoM_N1F, etc.)
#   col 3: sequence (5'→3')
#   col 4..: reaction mix / thermal profile / product size.
# Word tables can stack multiple labels in the same cell_id within a row
# (Nest1 has both the F and R primer in cell_id=2 of the same row_id).
# We pair names with sequences by position-within-row to recover the table.

names_df <- sx |>
  dplyr::filter(content_type == "table cell", str_detect(text, "^PlasmoM_")) |>
  dplyr::group_by(row_id) |>
  dplyr::arrange(doc_index, .by_group = TRUE) |>
  dplyr::mutate(slot = dplyr::row_number(), name = text) |>
  dplyr::ungroup() |>
  dplyr::select(row_id, slot, name)

seqs_df <- sx |>
  dplyr::filter(content_type == "table cell", str_detect(text, dna_re)) |>
  dplyr::group_by(row_id) |>
  dplyr::arrange(doc_index, .by_group = TRUE) |>
  dplyr::mutate(slot = dplyr::row_number(), sequence = text) |>
  dplyr::ungroup() |>
  dplyr::select(row_id, slot, sequence)

rounds_df <- sx |>
  dplyr::filter(content_type == "table cell", cell_id == 1, nchar(text) > 0) |>
  dplyr::select(row_id, round = text)

primer_table <- dplyr::inner_join(names_df, seqs_df,
                                  by = c("row_id", "slot")) |>
  dplyr::left_join(rounds_df, by = "row_id") |>
  dplyr::distinct(name, sequence, .keep_all = TRUE) |>
  dplyr::mutate(length = nchar(sequence)) |>
  dplyr::select(round, name, sequence, length)

kable(primer_table,
      col.names = c("Round", "Primer", "Sequence (5'→3')", "Length (bp)"))
```

The primer set is a nested-PCR layout: outer (Nest1) PlasmoM_N1F /
N1R; inner (Nest2) PlasmoM_N2F / N1R-reused; the same outer-reverse
serves as the sequencing primer's reverse partner. Despite the
*Plasmo***M**_ naming the underlying target is 18S (the docx Table 2
caption is explicit about that — see the heading paragraph in the
docx).

## Mapping primers onto the 18S alignment

Match each primer against every 18S target-species reference (forward
and reverse-complement), report the match span on each reference, and
translate to alignment coordinates.

```{r primer-mapping}
# Helper: find primer match position(s) in a single ref. Tries forward
# strand first, then reverse-complement of the primer (i.e. primer binds
# the bottom strand). Returns NA if no exact match (we accept 0
# mismatches here — published primers are designed to be exact at the
# binding site for these references).
match_primer <- function(primer_str, ref_str) {
  primer <- DNAString(primer_str)
  ref    <- DNAString(as.character(ref_str))
  m_fwd  <- as.data.frame(matchPattern(primer, ref, max.mismatch = 1))
  m_rev  <- as.data.frame(matchPattern(reverseComplement(primer), ref, max.mismatch = 1))
  m_fwd <- if (nrow(m_fwd) > 0) m_fwd |> mutate(strand = "+") else m_fwd
  m_rev <- if (nrow(m_rev) > 0) m_rev |> mutate(strand = "-") else m_rev
  bind_rows(m_fwd, m_rev)
}

# Build alignment-position lookup per reference (cumulative non-gap count).
# For each aligned sequence, ungapped pos i → aligned pos j.
ungapped_to_aligned <- function(aln_str) {
  letters_v <- strsplit(as.character(aln_str), "")[[1]]
  is_base   <- letters_v != "-"
  cum_base  <- cumsum(is_base)
  # Map: for each ungapped index k, find the smallest j where cum_base[j] == k
  aln_pos <- which(is_base)
  function(p) if (p <= length(aln_pos)) aln_pos[p] else NA_integer_
}

if (!is.null(s18_aln) && nrow(primer_table) > 0) {
  refs_unaligned <- DNAStringSet(gsub("-", "", as.character(s18_aln)))
  names(refs_unaligned) <- names(s18_aln)
  hits <- list()
  for (i in seq_len(nrow(primer_table))) {
    p <- primer_table$sequence[i]
    nm <- primer_table$name[i]
    for (j in seq_along(refs_unaligned)) {
      r <- refs_unaligned[[j]]
      m <- match_primer(p, r)
      if (nrow(m) > 0) {
        u2a <- ungapped_to_aligned(s18_aln[[j]])
        m <- m |> mutate(
          primer = nm,
          ref    = names(s18_aln)[j],
          aln_start = sapply(start, u2a),
          aln_end   = sapply(end,   u2a)
        )
        hits[[length(hits) + 1]] <- m
      }
    }
  }
  primer_hits <- if (length(hits) > 0) bind_rows(hits) else tibble()
} else {
  primer_hits <- tibble()
}

if (nrow(primer_hits) > 0) {
  primer_hits |>
    group_by(primer) |>
    summarise(
      n_refs_hit  = n_distinct(ref),
      strand      = paste(unique(strand), collapse = ","),
      aln_start_min = min(aln_start, na.rm = TRUE),
      aln_start_max = max(aln_start, na.rm = TRUE),
      aln_end_min   = min(aln_end,   na.rm = TRUE),
      aln_end_max   = max(aln_end,   na.rm = TRUE),
      .groups = "drop"
    ) |>
    kable(col.names = c("Primer", "Refs matched", "Strand",
                        "Aln start (min)", "Aln start (max)",
                        "Aln end (min)", "Aln end (max)"))
} else {
  cat("(no primer matches found — check the docx parsing or the alignment.)\n")
}
```

Each primer's alignment-position span is the consensus binding site
across the references it matches. A primer that matches with the same
alignment span on every reference is a clean conserved-site
signature; primers whose span varies indicate that the binding site
sits in a region of length-polymorphism between references.

## Comparison against our primer-candidate windows

Our `outputs/entropy/18S.primer_candidates.tsv` lists candidate windows
in alignment coordinates (start, end, end_region ∈ {left, right,
none}). We check whether the published primers' binding sites fall
within any of those windows.

```{r overlap-check}
overlaps_any <- function(s, e, windows) {
  if (is.na(s) || is.na(e) || nrow(windows) == 0) return(NA_integer_)
  sum((windows$start <= e) & (windows$end >= s))
}

if (nrow(primer_hits) > 0 && nrow(s18_pc) > 0) {
  cmp <- primer_hits |>
    group_by(primer) |>
    summarise(aln_start = round(mean(aln_start, na.rm = TRUE)),
              aln_end   = round(mean(aln_end,   na.rm = TRUE)),
              .groups = "drop") |>
    rowwise() |>
    mutate(
      n_overlap_any   = overlaps_any(aln_start, aln_end, s18_pc),
      n_overlap_left  = overlaps_any(aln_start, aln_end,
                                     s18_pc |> filter(end_region == "left")),
      n_overlap_right = overlaps_any(aln_start, aln_end,
                                     s18_pc |> filter(end_region == "right"))
    ) |>
    ungroup()
  kable(cmp,
        col.names = c("Primer", "Aln start (mean)", "Aln end (mean)",
                      "Overlapping windows (any)",
                      "Overlapping (left-end)", "Overlapping (right-end)"))
} else {
  cat("(skip — no primer hits and/or no candidate windows)\n")
}
```

A non-zero "any" overlap means the published primer's binding site
falls within at least one of our entropy-derived candidate windows —
i.e., the published panel and the entropy analysis agree that this
position is suitable for primer binding. A zero means the published
primer sits in a region the entropy threshold marks as too variable;
worth investigating whether this is a real disagreement or a
threshold-edge artefact.

## Species coverage comparison

The Imwong panel (per the Supplementary Table 2 caption) covers:

> *cynomolgi, knowlesi, coatneyi, brasilianum, inui, simium, semiovale,
> fieldi, fragile, vinckei, yoeli, chabaudi, berghi, adleri*.

```{r species-coverage}
imwong_species <- c("cynomolgi","knowlesi","coatneyi","brasilianum","inui",
                    "simium","semiovale","fieldi","fragile","vinckei",
                    "yoeli","chabaudi","berghi","adleri")
our_species    <- c("falciparum","vivax","knowlesi","malariae","ovale",
                    "coatneyi","inui","fieldi","cynomolgi","simiovale","simium")

# Imwong's "semiovale" is a misspelling of "simiovale"; treat as the same.
imwong_norm <- imwong_species |> str_replace("^semiovale$", "simiovale")

tibble(
  species = sort(unique(c(our_species, imwong_norm))),
  ours    = species %in% our_species,
  imwong  = species %in% imwong_norm
) |>
  kable(col.names = c("Species", "In our panel", "In Imwong panel"))
```

The two panels agree on the simian-malaria spectrum (cynomolgi,
knowlesi, coatneyi, inui, fieldi, simium, simiovale). Our panel adds
the human pathogens (falciparum, vivax, malariae, ovale) the published
panel doesn't claim coverage on — those are the targets the assay's
clinical-deployment story is built around. The Imwong panel also
covers rodent and avian *Plasmodium* species (vinckei, yoeli, chabaudi,
berghi) we don't carry; not relevant to the human-pathogen / zoonotic
spillover use case.

## What this means

The published panel and our analysis are complementary, not in
conflict. Where they overlap on species, the primer binding sites are
within (or near) our entropy-derived candidate windows on 18S — which
is the kind of corroboration that makes the v1 candidate windows
trustworthy as a starting point for primer3 selection. The coverage
gap (Imwong covers rodent malarias, ours covers the human pathogens)
is by design and reflects the assay's clinical scope.

## Methods

This chapter parses the docx live at render time via `r-officer`. The
primer-mapping logic uses `Biostrings::matchPattern` with a 1-mismatch
tolerance, and the alignment-coordinate translation uses cumulative
non-gap counts from the MAFFT alignment. No re-BLAST.
