---
title: "Chapter 7 — Methodological comparison with published assays"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(knitr)
  library(tidyverse)
})
```

Chapter 6 compared this assay against the Imwong et al. (2018) panel at
the primer level — same marker (18S), different windows on the same
alignment. This chapter zooms out to the methodological level and puts
the long-amplicon Nanopore design alongside the three published
*Plasmodium* speciation papers most relevant to the assay's
deployment claims: Lee et al. (2011) on macaque reservoirs and the
*P. knowlesi* phylogeny, Imwong et al. (2018) on ultrasensitive
nested-PCR detection of asymptomatic simian-malaria infections in
humans, and Swangsri et al. (2025) on SYBR-melt multiplex qPCR for
three simian species. Each represents a different point on the
sensitivity-resolution-cost surface, and each clarifies where the
v1 long-read panel adds value and where it does not.

## Three published assays, three eras of speciation

```{r three-paper-table}
tibble(
  Dimension = c(
    "Year",
    "Platform",
    "Markers",
    "Amplicon size",
    "Species detected",
    "Mixed infections",
    "Sensitivity",
    "Output",
    "Headline finding",
    "Methodological limitation"
  ),
  `Lee 2011 (PLoS Pathog)` = c(
    "2011",
    "Nested PCR + Sanger cloning",
    "18S (nested screen); full *csp* gene; whole ~6 kb mtDNA",
    "~6 kb (mtDNA); ~1.5–1.7 kb (csp)",
    "5 simian (knowlesi, inui, cynomolgi, coatneyi, fieldi)",
    "Yes — up to 6 mtDNA haplotypes per macaque infection",
    "Conventional nested PCR (~1–5 parasites/µL)",
    "Full sequence — alleles, haplotypes, phylogeny",
    "Macaques are *the* reservoir for *P. knowlesi*; no host-restricted lineages; TMRCA ~257 ka; population expansion 30–40 ka",
    "Labour-intensive cloning; mtDNA-only resolution is excellent *within* *P. knowlesi* but not designed as a multi-species panel"
  ),
  `Imwong 2018 (J Infect Dis)` = c(
    "2018/19",
    "Ultrasensitive nested qPCR (uPCR)",
    "18S rRNA (PlasmoM_N1F/N1R outer, N2F/N1R inner)",
    "~347 bp outer; ~310 bp inner",
    "knowlesi, cynomolgi confirmed (+ human spp via wider uPCR panel)",
    "Yes — detected dual *P. vivax* + *P. cynomolgi*",
    "LOD ~22 parasites/mL (best published for simian malaria)",
    "Species call + Sanger confirmation if positive",
    "Asymptomatic naturally-acquired *P. cynomolgi* and *P. knowlesi* infections occur in humans in Cambodia",
    "Sanger required for species call beyond the 18S genus screen; published primer binding sites sit in a highly conserved interior — sensitive but discards most discriminatory signal"
  ),
  `Swangsri 2025 (Sci Rep)` = c(
    "2025",
    "SYBR Green multiplex qPCR + melt curve",
    "*msp1*",
    "~100–200 bp per species",
    "knowlesi, cynomolgi, inui only",
    "Yes, in spiked DNA mixtures; capped by Tm spacing",
    "10 copies/µL",
    "Tm bin per species (85.2 / 78.0 / 82.5 °C)",
    "Cheap, fast, field-deployable three-species discrimination via distinct melt peaks",
    "Three-species ceiling — Tm space doesn't extend cleanly to more targets; can't distinguish strains within a species; cross-validated only against the four human-pathogen species, not against close simian relatives"
  )
) |>
  kable(align = "lll")
```

## Where the field sits today

There are essentially three molecular tiers for *Plasmodium*
speciation in routine use today.

The first tier is qPCR/melt or LAMP. Swangsri sits here, alongside
PET-PCR and several published probe-based qPCR panels. The platforms
are cheap, fast, and deployable in resource-limited settings, but each
new species costs a Tm slot or a probe channel and either runs out.
This tier is well-suited to surveillance triage where the question is
binary ("is this *P. knowlesi*, yes or no?") and the panel of interest
is small.

The second tier is nested PCR plus Sanger sequencing — the current
research workhorse and where most published macaque-survey literature
operates. Imwong's uPCR is the state of the art at this tier: a genus
screen on 18S followed by either species-specific nested primers or
direct sequencing of the inner amplicon. Sensitivity is excellent
(<50 parasites/mL) and the assay captures sequence, but it is slow,
single-marker, and the conserved 18S region that gives Imwong's
primers their broad amplification range is also why 18S alone cannot
discriminate every close pair — the *P. knowlesi* / *P. coatneyi* /
*P. cynomolgi* triangle is famously slippery on 18S.

The third tier is multi-marker phylogenetic typing — full *csp* plus
6 kb mtDNA in Lee 2011's hands. Gold-standard for resolution and the
only tier that recovers haplotype structure within mixed infections,
but not a routine assay: it required PCR-then-clone-then-Sanger of
multiple colonies per sample, and Lee's team spent the better part of
a year resolving 36 patients and 11 macaques. It does not scale to
hundreds of samples.

The honest methodological gap, for the simian and zoonotic species
this panel targets, is that no published assay simultaneously (a)
calls species reliably for all 11 targets, (b) resolves within-species
variants, (c) detects mixed infections at realistic abundance ratios,
and (d) is cheap enough to run on hundreds of samples. Imwong is
closest on sensitivity, Lee is closest on resolution, Swangsri is
closest on cost — none does all three.

## Where the long-read panel adds value

Setting the v1 design against the three published assays clarifies
where the methodological argument actually holds up.

**Coverage breadth.** Imwong's panel covers fourteen species but only
via genus screen plus Sanger confirmation; Swangsri's melt-curve
approach does not scale past three or four species cleanly. The v1
MIT + 18S long-amplicon panel is explicitly designed for eleven
species in a single workflow with sequence-level evidence per call,
rather than a Tm bin or a binary PCR result.

**Within-marker resolution.** Chapter 6 of this book showed that
Imwong's forward primers sit at 18S alignment positions 1683–1740 — a
58 bp region with mean entropy 0.011. That position is essentially
optimal for primer *binding* but contains almost no discriminatory
signal. By spanning 4–5 kb of 18S the v1 amplicons capture the
variable interior loops that actually separate species pairs. The same
logic applies to mtDNA: Lee used the full ~6 kb mtDNA for exactly this
reason — interior variation is where the species-pair signal lives,
and a short conserved amplicon throws most of it away.

**Mixed-infection resolution.** Lee's macaque samples carried up to
six mtDNA haplotypes per infection. Tm-based assays cannot see this;
Sanger-cloning is the painful way to recover it. Nanopore long reads
recover it natively because each read is a single template molecule,
so haplotype abundance comes for free with the basecalling.

**Cost in the right currency.** ~$5/sample at 96-plex is competitive
with nested qPCR per sample, and the workflow replaces "amplify →
clone → Sanger → align → species-call" (Lee/Imwong) with a single
sequencing run.

## Where it does not beat the existing assays

Sensitivity below ~100 parasites/µL is going to be a real challenge.
Imwong's uPCR will out-detect long-amplicon Nanopore on
low-parasitaemia asymptomatic carriers because 4–5 kb amplicons fall
off a cliff as template degrades or fragments. The v1 assay's natural
niche is symptomatic clinical samples and research-quality macaque
blood, not the low-density asymptomatic infections Imwong specialised
in. The framing in any deployment claim should be honest about this:
the long-read panel is a resolution tool, not a screen-everyone tool.

The other limitation worth flagging is that none of the v1 markers are
faster-evolving than 18S/mtDNA. If the assay needs to discriminate
within-species lineages (e.g. between geographic populations of
*P. knowlesi*) for epidemiological tracking — Lee's use case —
*csp* or *msp1* will outperform what MIT + 18S can deliver, regardless
of read length. The MSP1 expansion staged in
`scripts/R/msp1_full_gene_alignment.R` is partly a hedge against this:
if it improves per-pair resolution beyond MIT + 18S alone, the panel
gains a third marker; if it doesn't, the assay's published
specifications hold without modification.

## Sources

- Lee KS, Divis PCS, Zakaria SK, Matusop A, Julin RA, Conway DJ,
  Cox-Singh J, Singh B. *Plasmodium knowlesi*: Reservoir Hosts and
  Tracking the Emergence in Humans and Macaques. *PLoS Pathog*
  7(4):e1002015, 2011.
  [doi:10.1371/journal.ppat.1002015](https://doi.org/10.1371/journal.ppat.1002015)
- Imwong M, Madmanee W, Suwannasin K, Kunasol C, Peto TJ, Tripura R,
  von Seidlein L, Nguon C, Davoeung C, Day NPJ, Dondorp AM, White NJ.
  Asymptomatic Natural Human Infections With the Simian Malaria
  Parasites *Plasmodium cynomolgi* and *Plasmodium knowlesi*.
  *J Infect Dis* 219(5):695–702, 2019.
  [doi:10.1093/infdis/jiy519](https://doi.org/10.1093/infdis/jiy519)
- Swangsri T, Rucksaken R, Sricharern W, Nguitragool W, Saralamba N.
  Development and validation of a multiplex PCR assay with melt curve
  analysis for detecting simian *Plasmodium* in wild *Macaca
  fascicularis*. *Sci Rep* 15:22872, 2025.
  [doi:10.1038/s41598-025-07337-3](https://doi.org/10.1038/s41598-025-07337-3)
