---
title: "Chapter 8 — In-silico validation of the Swangsri 2025 MSP1 assay"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(Biostrings)
  library(knitr)
  library(tidyverse)
})

read_tsv_safe <- function(p) if (file.exists(p)) read_tsv(p, show_col_types = FALSE) else tibble()

fetch_log  <- read_tsv_safe("../data/reference/msp1/fetch_log.tsv")
strict     <- read_tsv_safe("../outputs/msp1/amplicon_specificity_strict.tsv")
lenient    <- read_tsv_safe("../outputs/msp1/amplicon_specificity_lenient.tsv")
ref_dir    <- "../data/reference/msp1"
```

Chapter 7 framed the methodological case for adding MSP1 as a third marker
and noted that the Swangsri et al. 2025 multiplex qPCR/melt assay covers
only three species (knowlesi, cynomolgi, inui). This chapter answers the
empirical follow-up: do their species-specific primers extend cleanly to
the eight panel species they didn't test, or do they cross-react / fail?

The check is in-silico PCR of all three Swangsri primer pairs against every
MSP1 reference for the 11-species panel that we could pull from GenBank.
Two mismatch tolerances are tested: strict (≤1 mismatch per primer — the
condition under which Swangsri's published cycling profile would actually
amplify) and lenient (≤3 mismatches — surfaces near-misses worth knowing
about).

## Reference coverage

The fetch (`scripts/python/fetch_msp1_refs.py`, `--max-seqs 200`) returns
records of widely varying length and completeness depending on what's
published per species. The summary below is what Phase A actually had to
work with.

```{r ref-coverage}
if (nrow(fetch_log) > 0) {
  # Length distribution per species — re-derived from the FASTAs because
  # the fetch log doesn't carry median length.
  ref_files <- list.files(ref_dir, pattern = "_msp1\\.fasta$",
                          full.names = TRUE)
  lens <- map_dfr(ref_files, function(f) {
    sp <- sub("_msp1\\.fasta$", "", basename(f))
    s  <- tryCatch(readDNAStringSet(f), error = function(e) NULL)
    if (is.null(s) || length(s) == 0)
      return(tibble(species = sp, n_refs = 0L,
                    median_len = NA_integer_, n_full_cds = 0L))
    L <- width(s)
    tibble(species    = sp,
           n_refs     = length(s),
           median_len = as.integer(median(L)),
           n_full_cds = sum(L >= 5000L))
  })
  full_panel <- c("falciparum","vivax","knowlesi","malariae","ovale",
                  "coatneyi","inui","fieldi","cynomolgi","simiovale","simium")
  out <- tibble(species = full_panel) |>
    left_join(lens, by = "species") |>
    mutate(across(c(n_refs, n_full_cds),
                  ~ replace_na(.x, 0L)))
  kable(out,
        col.names = c("Species", "Refs kept", "Median length (bp)",
                      "Full-CDS (≥5 kb)"),
        align = "lrrr")
} else {
  cat("(no fetch log — run `python3 scripts/python/fetch_msp1_refs.py` first.)\n")
}
```

Two species are untestable in-silico: *P. simium* returns zero MSP1
records on GenBank under the query, and the single *P. simiovale* record
returned did not pass the species-name normaliser. Both need wet-lab
confirmation or manually curated references before any conclusion can be
extended to them.

The records dominating the falciparum, vivax, knowlesi, malariae and ovale
panels are short 3'-end CDS fragments from vaccine and diversity studies.
The Swangsri primers bind near the 5' end of msp1 (~600–820 bp from gene
start), so these partial records don't contain the primer-binding region
and would never amplify regardless of primer specificity. The
`Full-CDS (≥5 kb)` column is what matters for interpreting the next two
tables — it's the subset of refs that actually span the binding window.

## On-target sanity check

Before believing any cross-species result, the on-target case has to work
at strict tolerance — i.e., each Swangsri primer pair must amplify its
nominal target species cleanly. If it doesn't, the primer sequences in
the script are wrong or the records are mis-annotated, and the
cross-species table is meaningless.

```{r on-target}
if (nrow(strict) > 0) {
  on_target <- strict |>
    filter(on_target) |>
    group_by(primer_pair, target_species) |>
    summarise(
      refs_tested     = n(),
      refs_amplifying = sum(!is.na(amplicon_len)),
      best_fwd_mm     = suppressWarnings(min(fwd_mismatches, na.rm = TRUE)),
      best_rev_mm     = suppressWarnings(min(rev_mismatches, na.rm = TRUE)),
      median_amp_len  = suppressWarnings(median(amplicon_len, na.rm = TRUE)),
      .groups         = "drop"
    ) |>
    mutate(across(starts_with("best_"),
                  ~ ifelse(is.infinite(.x), NA_integer_, .x)),
           median_amp_len = ifelse(is.na(median_amp_len),
                                   NA_integer_, as.integer(median_amp_len)))
  kable(on_target,
        col.names = c("Primer pair", "Target", "Refs tested",
                      "Refs amplifying", "Best fwd mm", "Best rev mm",
                      "Median amp len (bp)"),
        align = "llrrrrr")
}
```

All three on-target pairs amplify at 0/0 mismatches on at least one
reference. **Sanity check passes** — the cross-species result is
trustworthy.

The low refs-amplifying counts on knowlesi and inui are driven by the
partial-record bias noted above: most of those refs simply don't span the
5' binding window. Cynomolgi is the outlier, treated separately below.

## Cross-species specificity

The headline question. Do any of the three primer pairs amplify off-target
species?

```{r cross-species}
if (nrow(strict) > 0 && nrow(lenient) > 0) {
  off_strict <- strict |>
    filter(!on_target) |>
    group_by(primer_pair, target_species) |>
    summarise(refs_tested        = n(),
              refs_amplifying_s  = sum(!is.na(amplicon_len)),
              .groups = "drop")
  off_lenient <- lenient |>
    filter(!on_target) |>
    group_by(primer_pair, target_species) |>
    summarise(refs_amplifying_l  = sum(!is.na(amplicon_len)),
              best_fwd_mm_l      = suppressWarnings(min(fwd_mismatches, na.rm = TRUE)),
              best_rev_mm_l      = suppressWarnings(min(rev_mismatches, na.rm = TRUE)),
              .groups = "drop")
  off <- off_strict |>
    left_join(off_lenient, by = c("primer_pair", "target_species")) |>
    mutate(across(starts_with("best_"),
                  ~ ifelse(is.infinite(.x), NA_integer_, .x))) |>
    arrange(primer_pair, target_species)
  kable(off,
        col.names = c("Primer pair", "Off-target species", "Refs",
                      "Amp @ strict (≤1mm)", "Amp @ lenient (≤3mm)",
                      "Best fwd mm (lenient)", "Best rev mm (lenient)"),
        align = "llrrrrr")
}
```

**Zero productive off-target amplicons** at either tolerance, against any
of the eight non-target species with reference data. The three near-miss
patterns at lenient (Pcy forward ≤1 mm vs fieldi; Pk reverse ≤3 mm vs
coatneyi; Pk reverse ≤3 mm vs inui) all have the partner primer with no
binding-region match anywhere on those references — so neither half-bind
can produce a product under realistic PCR cycling. The Swangsri assay is
genuinely species-specific against the panel.

## The within-species cynomolgi concern

The on-target table flags a real issue distinct from the cross-species
question Phase A was designed to answer. Pcy amplifies only 1 of 25
cynomolgi references at strict tolerance, rising to 2 of 25 at lenient.

This is not the partial-record artefact that depresses the knowlesi and
inui on-target counts. The cynomolgi panel has 13 of 25 records as
complete CDS (median length 5309 bp), so the binding region *is* present
in most refs. The reverse primer site is the bottleneck: across the
full-CDS cynomolgi records that do contain the binding region, most carry
≥ 2 mismatches in the Pcy_MSP1_R699 footprint.

The one record that amplifies is **AB266195.1, strain "Pt2 (Berok)"** —
the same accession Swangsri used as the single cynomolgi sequence in
their Methods Fig. 1 alignment, and the basis for their primer design.
The other named cynomolgi strains in the GenBank panel — *Mulligan*
(AB266193.1), *B* (AB266188.1), *Ceylonensis* (AB266190.1), *Cambodian*
(AB266189.1), *Langur*, *Gombak*, *Smithsonian*, *T824*, *T746*, plus
strain *ATCC 30146* (AB444063.1) — none of them amplify at strict
tolerance.

The Pt2/Berok strain is the outlier in the cynomolgi panel, not the
representative.

```{r pcy-cynomolgi-detail}
if (nrow(lenient) > 0) {
  pcy_cyn <- lenient |>
    filter(primer_pair == "Pcy", target_species == "cynomolgi") |>
    summarise(
      n_refs                 = n(),
      n_fwd_binds_lenient    = sum(!is.na(fwd_mismatches)),
      n_rev_binds_lenient    = sum(!is.na(rev_mismatches)),
      median_fwd_mm          = suppressWarnings(median(fwd_mismatches, na.rm = TRUE)),
      median_rev_mm          = suppressWarnings(median(rev_mismatches, na.rm = TRUE)),
      n_productive_amplicons = sum(!is.na(amplicon_len))
    )
  kable(pcy_cyn,
        col.names = c("Cynomolgi refs", "Fwd binds (≤3mm)",
                      "Rev binds (≤3mm)", "Median fwd mm",
                      "Median rev mm", "Productive amplicons"),
        align = "rrrrrr")
}
```

The asymmetry between forward-primer binding and reverse-primer binding,
combined with the median mismatch count on the reverse site, is the
signal. The forward site is conserved across cynomolgi isolates; the
reverse site isn't.

### How the published validation could have missed this

Two design decisions in the Swangsri paper compound to make this
invisible to the validation pipeline they ran.

**The design reference is not the lab strain.** Swangsri designed the
Pcy primers manually in BioEdit against AB266195.1 — the single
cynomolgi sequence in their Methods Fig. 1 alignment. AB266195.1 is
strain Pt2 (Berok), a Malaysian field isolate. Their lab work, per
Methods, used genomic DNA purchased from ATCC, which corresponds to
strain ATCC 30146 (AB444063.1 in GenBank). These are different strains
with different sequence at the Pcy reverse primer site. The in-silico
check predicts no productive amplicon from AB444063.1 even at 3-mismatch
tolerance — yet Swangsri's wet-lab PCR succeeded in amplifying the
binding region from ATCC genomic DNA to make their plasmid clone. The
most likely reconciliation is that their qPCR cycling conditions are
permissive enough to amplify through the mismatches against ATCC at
reduced efficiency, particularly because the mismatches in the forward
primer footprint sit away from the 3' end where they would be most
disruptive.

**The validation was done against the plasmid clone, not the strain.**
The plasmid clone is a copy of the binding region as it exists in the
specific ATCC stock they sequenced, made by cloning a PCR product into
pGEM-T. Once the plasmid exists, it is by construction a perfect match
to the primer sites — even if the original ATCC genomic template
required mismatches to amplify in the cloning step. The sensitivity,
LOD, reproducibility, and efficiency assays in the paper all used
dilutions of this plasmid; they measure how well the assay performs
against a sequence-perfect template, not against natural strain
diversity.

**The field arm had no test of cynomolgi-against-natural-isolates.**
The study screened 191 wild *M. fascicularis* samples from Thailand and
found zero *P. cynomolgi* positives — only two *P. knowlesi*. The
empirical arm where natural cynomolgi strain diversity would have
surfaced had no positive samples to test.

The result is a validation pipeline that is internally consistent — the
primers amplify a clone of the binding region as it exists in one
specific ATCC stock, and they don't cross-react with the four human
*Plasmodium* species. Both of those claims are defensible. But neither
addresses the question of whether the Pcy primers amplify natural
cynomolgi infections in the field, because no step in the validation
chain involves natural cynomolgi templates.

This is not unique to Swangsri — it's a structural feature of
plasmid-based qPCR validation against single-strain lab cultures, and
it applies to many published *Plasmodium* qPCR assays. Surfacing
within-species primer-binding polymorphism requires either (a) an
alignment of multiple natural isolates of the target species (which
this analysis provides, in silico) or (b) running the assay against a
panel of natural-infection samples covering the species' diversity.

## Headline and recommendation

The Swangsri 2025 MSP1 species-specific assay extends cleanly to the
panel species they did not test. No predicted off-target amplification
for any of the three primer pairs against any of the eight non-target
species with available MSP1 reference data, at strict or lenient
tolerance. Two species (*P. simium*, *P. simiovale*) remain untestable
in-silico and need wet-lab confirmation.

The actionable finding for colleagues validating the assay is the within-
species cynomolgi sensitivity concern: 1–2 of 25 cynomolgi references
amplify under predicted conditions, driven by polymorphism in the
reverse-primer site. This is independent of the cross-species question
and is the most likely failure mode of the assay in real-world use on
natural cynomolgi infections.

A defensible recommendation to the Swangsri authors and to anyone
considering deploying the assay: redesign or supplement the Pcy reverse
primer using an alignment of natural cynomolgi MSP1 isolates rather than
a single reference, and re-validate against multiple natural cynomolgi
isolates (not just the cloned plasmid) before claiming species-level
sensitivity for cynomolgi detection.

## Methods

In-silico PCR uses `Biostrings::vmatchPattern` with configurable mismatch
tolerance. For each reference, both primers are tested against both
strands; a productive amplicon requires the forward primer on the +
strand 5' of the reverse primer on the − strand, with an inter-primer
distance < 2000 bp. Predicted Tm proxy is the Wallace-rule sum
(4×[G+C] + 2×[A+T]) and is reported only for context — it is not directly
comparable to SYBR-melt qPCR Tm values and was not used to draw any
conclusion in this chapter. Source: `scripts/R/msp1_amplicon_specificity.R`.

References were fetched via `scripts/python/fetch_msp1_refs.py`
(NCBI E-utilities, `--max-seqs 200` per species) with FASTA headers
normalised through the project's case_when species ladder. The
`--max-seqs 200` setting was needed to surface full-CDS records — the
default 20 returns mostly partial 3'-end fragments that don't span the
Swangsri primer-binding region.
